Estimating dark matter halo concentration using the integrated mass profile

### 2015-11-09 paper/paper.tex
ahora la intro fluye mejor

Galaxy properties are driven by its dark matter halo host.
A dark matter halo's concentration is on the key structural
properties. 
[EN QUE MODELOS IMPORTA LA CONCENTRACION?]
This drives the pursue to make the concentration estimates of
simulated halos as reliable as possible. 
Here we show the radially integrated mass profile can be converted
into a powerful tool to estimate concentratios.

The main advantage of this approach is that it uses the full
particle information without any binning, which makes it reliable in
cases when low numerical resolution becomes a limitation for other
methods.   
We show the good performance of this method by estimating halo
concentration both on mock and N-body halos. 
We compare these results against two other methods:
maximum radial velocity measurements and radial particle binning to
fit the density profile. 
Tests on  the mock halos show that the accuracy of our method to recover input
concentrations varies with halo resolution. For halos
sampled with $20$ particles our method recovers the input
concentration with $10\%$ accuracy. 
For the maximum radial velocity method it is on the order of $20\%$
while density profile methods perform worst at the low resolution
limit since they require at least 200 particles inside the virial
radius to make a reliable estimate of the density profile. 
For halos with $2\times 10^4$ particles our method achieves an
accuracy of $0.01\%$ while the velocity and density profile methods
achieve $0.1\%$ and $1\%$, respectively. 
We measure the mass-concentration relationship on N-body
data taking care of using halos sampled with at least $200$
particles. We find that at low masses $10^{12}\hMsun <
M<10^{13}\hMsun$ our method yields median concentration values lower
by a $15\%-20\%$ compared to the velocity and density methods ({\bf
  why?}).  
At higher masses $3\times 10^{13}\hMsun < M < 2\times
10^{14}\hMsun$ the three methods give similar results.
\end{abstract}

@@

%*************************************************************************
\section{Introduction}
\label{sec:introduction}
%*************************************************************************
In the concordance cosmological paradigm the matter content of the
Universe is dominated by dark matter, a collisionless fluid shaped by
gravitational interactions.  In the last three decades simulations of
dark matter dominated universes have provided valuable insights into
the large scale structure formation process, showing a remarkable
success in the comparison between theoretical results and observations
of the galaxy distribution obtained from surveys.
\citep{Springel2005,2011ApJ...740..102K}.

{\bf FALTA CITEP Navarro et.al. 2010} have used high resolution simulations to show
that the full halo density profile is not universal and follows an
Einasto density distribution. 
However, if one is not interested in the very central region of the
dark matter halo (where galaxy formation takes place, and where the
effects of baryon physics on dark matter distribution is still an
uncknown), one can assume an approximation where dark matter
overdensities closely follow an universal density profile. 
In a first approximation this profile is spherically symmetric and
its density only dependens on the radial coordinate. 
Although it is well known that the characteristics of
this densty profile depend on redshift {\bf FALTA CITEP (Maccio
  et.al. 2008)}, its universal shape seems to be independent of the
cosmological parameters and is self-similar for different spatial
scales after an adequate re-scaling is applied \citep{NFW,Taylor2001}.

In such circumstances, one the most popular parameterizations
for dark matter halo radial density distribution is the
Navarro-Frenk-White (NFW) profile \citep{NFW}.  
This profile is a double power law in radius, where the transition
break happens at the so-called scale radius, $r_s$.  
The ratio between the scale radius and the virial radius $R_v$ is
known as the concentration $c=R_v/r_s$. 
Simulations show that the concentration is a strong function of halo
mass and redshift and provides information about the mass assembly of
the dark matter halo {\bf (Citar paper que reproduce mis graficas y conecta
C con m(z))}.

{\bf usos de modelos de formacion de galaxias}
The relationship between halo mass and concentration
could also provide a potential test of the Cold Dark Matter (CDM)
paradigm.

For all these reasons mentioned above a great deal of effort has
been invested in calibrating this relationship with simulations
\citep{Neto2007,Duffy2008,Munoz2011,Prada2012,Ludlow2014} and finding
the best possible way to constraint it with observations
\citep{Buote2007,Comerford2007,Mandelbaum2008,Giocoli2014,Foex2014,Shan2015}.

In N-body simulations there are two main methods to estimate the
concentration parameter of a dark matter halo.
The first method takes the halo particles and bins them into
logarithmic radii to estimate the density in each bin, then it 
proceeds to fit the density as a function of the radius to the NFW
profile.  
A second method uses an analytic property of the NFW profile
that relates the maximum of the ratio of the circular velocity to the
virial velocity.  The concentration can be then found as the root of a
algebraic equation dependent on this maximum value.

@@

The second method solves the two problems mentioned above.  
It works with low particle numbers and does not involve data binning.  
However, it effectively takes into account only a single data point and
discards the behaviour of the ratio $V_{\rm circ}/V_{\rm vir}$ below
and above its maxima.  
Additionally, small fluctuations on the value of this maximum can
yield large perturbations on the estimated concentration parameter.

In this paper we show that integrated mass profile can be used  to
estimate the dark matter halo concentration in halos from N-body
simulations.  
This has two advantages with respect to the two methods mentioned
above.  
It does not involve any data binning and does not throw away data
points resulting in a robust estimate even at low resolution/particle
numbers.  
Furthermore, since the method does not require any binning,
there is no need to tune numerical parameters, providing a new
independent method to estimate the concentration parameter.

>>>>

Galaxy properties are coupled to the structural of its dark
matter halo host. 
One of the key quantities is the halo density profile which, assuming
a  Navarro-Frenk-White parameterization, can be described by the
concentration.
This has motivated the development of different algorithms to estimate
the concentraion in N-body halos.
Here we present a new algorithm to estimate the concentration
based on the radially integrated mass profile
The main advantage of this approach is that it uses the full
particle information without any binning, which makes it reliable in
cases when low numerical resolution becomes a limitation for other
methods.   
We show the good performance of this method by estimating halo
concentration both on mock and N-body halos. 
We compare these results against two other methods:
maximum radial velocity measurements and radial particle binning to
fit the density profile. 
Tests on  the mock halos show that the accuracy of our method to recover input
concentrations varies with halo resolution. For halos
sampled with $20$ particles our method recovers the input
concentration with $10\%$ accuracy. 
For the maximum radial velocity method it is on the order of $20\%$
while density profile methods perform worst at the low resolution.
For halos with $2\times 10^4$ particles our method achieves an
accuracy of $0.01\%$ while the velocity and density profile methods
achieve $0.1\%$ and $1\%$, respectively. 
We measure the mass-concentration relationship on N-body
data taking care of using halos sampled with at least $200$
particles. We find that at low masses $10^{12}\hMsun <
M<10^{13}\hMsun$ our method yields median concentration values lower
by a $15\%-20\%$ compared to the velocity and density methods 
At higher masses $3\times 10^{13}\hMsun < M < 2\times
10^{14}\hMsun$ the three methods give similar results.
\end{abstract}

@@

%*************************************************************************
\section{Introduction}
\label{sec:introduction}
%*************************************************************************
In the current structure formation paradigm the properties of galaxies
are coupled to the evolution of their dark matter (DM) hosting halo.
This dependence has motivated the detailed study of dark matter halo
properties defining its formation history and internal structure. 
In particular, the DM distribution inside a halo is key to
understand the sizes and dynamics of galaxies.

The internal matter distribution is parameterized through the density
profile. 
In a first approximation this profile is spherically symmetric and
its density only dependens on the radial coordinate. 
One the most popular parameterizations for DM halo radial
density distribution is the Navarro-Frenk-White (NFW) profile
\citep{NFW}.   
Furthermore, if one is not interested in the very central region of the
dark matter halo (where galaxy formation takes place, and where the
effects of baryon physics on dark matter distribution is still an
unknown), one can assume that this profile is universal.
\citep{Navarro2010}.  
 This profile is a double power law in radius, where the transition
break happens at the so-called scale radius, $r_s$.  
The ratio between the scale radius and the virial radius $R_v$ is
known as the concentration $c=R_v/r_s$.

The NFW profile parameterization thus provides the conceptual
framework to study simulated dark matter halos as a function of
redshift and cosmological parameters.
Here the central quantity is the concentration. 
In these studies 
\citep{Neto2007,Maccio2008,Duffy2008,Munoz2011,Prada2012,Ludlow2014} 
the results are summarized through the
mass-concentration relationship; that is the distribution of
concentration values at a fixed halo mass and redshift.
The biggest challenge in such numerical experiments is having a
reliable algorithm to estimate the concentration. 
Such an algorithm should provide unbiased results and must be robust
when applied to simulations of different numerical resolution.

There are two main algorithms to estimate the concentration parameter
of a N-body dark matter halo. 
The first method takes the halo particles and bins them into
logarithmic radii to estimate the density in each bin, then it 
proceeds to fit the density as a function of the radius to the NFW
profile.  
A second method uses an analytic property of the NFW profile
that relates the maximum of the ratio of the circular velocity to the
virial velocity.  The concentration can be then found as the root of a
algebraic equation dependent on this maximum value.

@@

The second method solves the two problems mentioned above.  
It works with low particle numbers and does not involve data binning.  
However, it effectively takes into account only a single data point and
discards the behaviour of the ratio $V_{\rm circ}/V_{\rm vir}$ below
and above its maxima.  
Small fluctuations on the value of this maximum can yield large
perturbations on the estimated concentration parameter.

In this paper we present a third alternative based on fitting the
integrated mass profile.
This approach has two advantages with respect to the above mentioned
methods.  
It does not involve any data binning and does not throw away data
points. 
This results in a robust estimate even at low resolution/particle
numbers.   
Furthermore, since the method does not require any binning, there is
no need to tune numerical parameters, providing a new independent
method to estimate the concentration parameter. 
Here we provide a detailed description of this method and a
comparison against the two traditional methods by measuring its
performance both on mock halos and N-body halos.
